Integration of national demographic-disturbance relationships into local caribou population viability projections can reduce uncertainty and inform monitoring decisions

### 2024-10-15 analysis/paper/paper.Rmd
revisions in progress

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country. Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and reduce the risk that lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a term derived from a simple model of the recruitment survey observation process.  We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years may be better than intensive short-term monitoring for improving the reliability of status projections. 
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

The number of species threatened with extinction is high and getting
higher [@ipbes_global_2019]. Reliable information on the status of
populations and likely impacts of management are needed to inform
conservation action but effective monitoring is challenging and costly
[@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;
@dunham_strategic_2023]. Resources to support conservation and recovery
are limited [@buxton_balancing_2022; @mccarthy_financial_2012;
@gerber_conservation_2016; @wintle_spending_2019], delays increase risk
[@martin_acting_2012], and some populations have been monitored to
extinction [@lindenmayer_counting_2013]. A recent analysis of action
plans for Canadian species at risk found half of tasks are research and
monitoring, and this proportion is highest for species with higher risk
of extinction, highlighting a need to more carefully consider when,
where, and why additional information is needed to inform conservation
action [@buxton_balancing_2022]. Analysis of the implications of
monitoring outcomes for decisions can help clarify whether additional
monitoring is necessary to inform conservation action
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012].

Analysis methods that can help make better use of available monitoring
data to inform decisions include Bayesian population modelling
[@schaub_integrated_2021] and value of information analyses
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012]. To be useful, these methods must be tailored
to particular species and decision contexts. Here, we offer a simple
Bayesian population model for boreal caribou (*Rangifer tarandus
caribou*) that integrates information about the impacts of landscape 
change [@johnson_science_2020] with available demographic data [building on @eacker_web-based_2019] to reduce uncertainty in
population viability projections. Following the value of information analysis
framework proposed by Dunham et al. (2023), we combine this Bayesian
model with simulations of plausible population trajectories
[@stewart_climate-informed_2023; @dyson_existing_2022] to show how the
need for local monitoring varies with landscape condition, and to assess
the ability of alternative monitoring scenarios to reduce the risk of
errors in population viability projections. Our approach is tailored to
boreal caribou, but could be adapted for other species at risk found in
multiple populations that are costly to monitor and sensitive to habitat
disturbance.

Boreal caribou are widely and sparsely distributed through the Canadian
boreal region across a gradient of anthropogenic disturbance
[@eccc_scientific_2011; @johnson_science_2020] that includes some of the
most intact and inaccessible places remaining on earth
[@hirsh-pearson_canadas_2022; @ibisch_global_2016;
@venter_sixteen_2016]. Across 58 boreal caribou study areas, both
survival and recruitment decrease with the percentage of the study area
that is disturbed, and no populations of caribou are considered
self-sustaining in study areas with high anthropogenic disturbance
[@eccc_scientific_2011; @johnson_science_2020]. The causes of these
demographic-disturbance relationships include direct impacts on habitat
availability and use, and indirect impacts on predators and alternate
prey [@nbckc_boreal_2022; @eccc_scientific_2011]. Decreases in
recruitment as disturbance increases are also observed over time in
places where there is sufficient historic demographic data
[@rudolph_demographic_2017; @hervieux_widespread_2013], but many places
lack sufficient data to characterize local demographic-disturbance
relationships [@johnson_science_2020]. There is also variation in
demographic rates among study areas, particularly where anthropogenic
disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
@fortin_forest_2017; @gagne_impact_2016], and this variation leads to
uncertainty in projections derived from national disturbance-demographic
relationships [@stewart_climate-informed_2023; @dyson_existing_2022].

Interventions such as wolf culls, maternal penning, and restoration of
linear features are intended to improve survival and reproduction of
populations that are not otherwise self-sustaining
[@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a
need to assess and compare effects of boreal caribou conservation
actions with careful monitoring. Our focus in this paper is assessing
and projecting self-sustainability of populations in the absence of
these interventions, to clarify when additional information is needed to
reduce uncertainty about whether there is a need for action. Where
actions are taken, outcomes should be monitored; the simple model we
propose here does not include effects of management interventions, and
cannot be used to project outcomes for managed populations.

@@

Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, improve viability projections on changing
landscapes, and help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach can help clarify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

@@

To derive a composition survey bias term we assume each group
of animals in a calf:cow composition survey contains one or more
collared adult females ($T$), and may also include: uncollared adult
females misidentified as young bulls or unknown sex ($U$); correctly
identified uncollared adult females ($V$); young bulls correctly
identified as male or unknown sex ($O$); young bulls misidentified as
uncollared adult females ($P$); observed calves ($J$); and unobserved
calves ($K$). The apparent number of adult females in the group is
$T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
uncollared adult females to collared animals in the recruitment survey.
The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying
young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and
$P=(O+P)u$. Given a probability $z$ of missing calves, we get
$J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
bias-corrected recruitment rate $X$ can be written as a function of the
observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
bulls to adult females $q$, and the misidentification probabilities $u$
and $z$ (see Supplement B for details). For simplicity, we combine the
probabilities into a bias term $c$ that is integrated into
the Bayesian population model and simulations (below):
$$c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2$$

@@

In simulated monitoring scenarios, we assume collars are deployed in January, and
last for up to 6 years. The target number of collared females $n$ varies
among scenarios, as does the total duration of the monitoring program
$d$ and the ratio of uncollared to collared adults in the recruitment
survey $w$ (Table \@ref(tab:monitoringScenarios)). Collars lost to
mortality may be replaced each year ($o=1$), or there may be a 3 year
gap between deployments ($o=4$).

@@

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations (Figure
\@ref(fig:exampleTrajectories)). To ensure results are comparable across
monitoring scenarios and that populations remain extant through the
monitoring period, all simulations are started with $N_0=5000$ animals
24 years prior to the start of the projection period. Realized population growth rate is $\dot{\lambda}_t=\dot{N}_t/\dot{N}_{t-1}$ Expected population growth rate (without interannual variation, demographic stochasticity, or effects of population density) is $\bar{\lambda}_t=\bar{S}_t(1+\dot{c}\bar{R}_t/2)$.

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships. To highlight how variation among trajectories persists as disturbance changes we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$). Orange dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models.", cache=TRUE}

@@

For consistency with Johnson et al. [-@johnson_science_2020], we model survival
probability as an adjusted function of buffered anthropogenic
disturbance $A_t$ with a log link and Beta distributed interannual variation with coefficient of variation $\nu_S$, mean $\tilde{S_t}$ and standard deviation $\sigma_{S,t}$ [@stewart_climate-informed_2023;
@dyson_existing_2022]: 
\begin{equation}
\tilde{S_t}=(46 e^{\beta^S_0+\beta^S_a A_t}-0.5)/45
  (\#eq:Stilde)
\end{equation}

@@

For reference, all variables and parameters are defined in supplemental Table A1. Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chain,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figures
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
uncertainty about the population growth rate decreases as sampling
effort in each year (i.e. number of collars) or the number of years of
monitoring (Figure \@ref(fig:figDiffs)) increase. Uncertainty is higher
when the forecast horizon is long (compare 5 and 20 year projections in
Figure \@ref(fig:figDiffs)), and when interannual variation is high (Figure
\@ref(fig:examplePredictions)). Uncertainty is also higher in our high
anthropogenic disturbance scenarios (Figure \@ref(fig:figDiffs)) because
stochastic variation in population growth rates increase as populations
become very small (as in top right panel of Figure
\@ref(fig:examplePredictions)).

@@

The availability of boreal caribou demographic data varies widely across
Canada due to large variation in habitat condition, population
condition, available resources for monitoring, and remoteness. In a few
regions enough data has been collected over time to characterize local
demographic-disturbance relationships [@rudolph_demographic_2017;
@hervieux_widespread_2013], but most populations have only been sparsely
monitored for snapshot estimates of trend or abundance
[@johnson_science_2020]. As a result, in most regions it is necessary to
make assumptions informed by observed variation in demographic rates
with disturbance among populations and other relevant information about
impacts of disturbance on caribou. However, high observed variation in
demography where anthropogenic disturbance is low leads to high
uncertainty in projections informed by national demographic-disturbance
relationships alone [@stewart_climate-informed_2023;
@dyson_existing_2022]. In some cases, predictions of population growth
rate informed by national demographic-disturbance relationships appear
to disagree with estimates from local demographic data
[@eccc_scientific_2011; @dyson_existing_2022], but without formal
methods of accounting for uncertainty it can be difficult to distinguish
true discrepancies from stochastic variation.

We propose a simple Bayesian population model that helps address these challenges by
acknowledging uncertainty associated with local demographic data and
national demographic-disturbance relationships, and integrating these
two sources of information to reduce uncertainty in projections of the
impacts of changing landscape conditions. This Bayesian approach
represents an important improvement over less formally integrated
methods for estimating and projecting impacts of disturbance on
population population viability [@eccc_scientific_2011;
@johnson_science_2020] because it reconciles and accounts for
uncertainties in local demographic data and demographic-disturbance
relationships.

Our model assumes that increasing cumulative anthropogenic disturbance
causes a log-linear decrease in caribou demographic rates (equations 4,
5, 10 and 13), reflecting a scientific consensus informed by a number of
lines of evidence that cumulative anthropogenic disturbance has a
negative impact on boreal caribou [@eccc_scientific_2011;
@johnson_science_2020; @gouvernement_du_quebec_revue_2021;
@rudolph_demographic_2017; @sorensen_determining_2008;
@hervieux_widespread_2013]. Local data can be integrated to estimate
local demographic-disturbance parameters that differ from the national
means, and in the absence of local data, appropriate priors lead to a
range of outcomes consistent with the observed range of variation across
the country [Figures 3 and 5 in @johnson_science_2020]. Others
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @dalgarno_bboutools_2024] have also used Bayesian models to estimate
caribou demographic rates from local monitoring data, but our addition
of prior information from national demographic-disturbance relationships
is novel, and important for reducing the risk that lack of local data
will be used as a reason to delay conservation action for highly
disturbed and poorly monitored populations. Integrating
demographic-disturbance relationships into a viability projection model
clarifies that no additional information is needed to project outcomes
where disturbance is high and actions to support recovery are not taken.

@@

Variation in anthropogenic disturbance across the Canadian boreal region
is partially correlated with variation in climate and primary
productivity [@neilson_direct_2022] because forestry occurs in higher
productivity areas, and petroleum extraction activities are concentrated
in Western Canada. Effects of disturbance on vegetation and foodweb
dynamics vary with primary productivity [@neufeld_primary_2021], and
some have expressed concern about confounding of demographic-disturbance
relationships [@wilson_causal_2021]. One option for addressing these
concerns would be to develop regional demographic-disturbance
relationships for populations that are ecologically similar to one
another (e.g. caribou inhabiting forests that are sufficiently
productive for industrial forestry in Eastern Canada)
[@eccc_scientific_2024], and use the priors from regional
demographic-disturbance relationships to inform demographic projections
within the region. However, while acknowledging and allowing for
regional variation, we also need to guard against risks that uncertainty
will be inflated and used as an excuse to delay action, a tactic that
has been used in this and other decision contexts
[@bissonnette_controverses_2020; @boan_climate_2018]. In order to be
informative and credible, regional demographic-disturbance analyses must
include a sufficient number of observations spanning a range from high
to low anthropogenic disturbance. Priors that lead to projections of
outcomes for caribou that fall outside the observed range of variation
across the country, characterized here (Figure 2) by 95% prediction
intervals from a national analysis (Johnson et al. 2020), are
implausible. In other words, if we have seen no examples of
self-sustaining caribou populations at a particularly high level of
disturbance, it is unreasonable to project or expect this outcome. These
premises constrain the plausible demographic-disturbance priors, and
clarify that no additional information is needed to predict outcomes for
caribou where anthropogenic disturbance is high (Figure 5) and no
meaningful actions are taken to support recovery.

@@

Another novel aspect of this work is a term that acknowledges possible bias
caused by misclassification of adults and missing calves in
composition surveys [@ellington_divergent_2020;
@decesare_estimating_2012]. Lacking information about the true magnitude
of these errors, we examined sensitivity to a range of assumed values, and
found that acknowledging bias did increase uncertainty, but these
increases were not large (Supplement E). A key reason for this result is
that we assumed the same range of values for the probability of
misclassifying adults and the probability of missing calves. Presuming
that surveyed groups contain more cows than bulls, and that there is an
equal probability of misclassifying cows and young bulls and vice versa,
then misclassification errors cause high estimated recruitment. Missing
calves causes low estimated recruitment, and together these errors tend
to cancel one another out. Given lack of knowledge about true composition survey error rates, the most appropriate interpretation of our results is that better information is required in order to assess the importance of these errors.

>>>>

institute:
  - ECCC: Landscape Science and Technology Division, National Wildlife Research Center, Environment and Climate Change Canada, 1125 Colonel By Dr, Ottawa, ON, K1S 5B6, Canada
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      number_sections: no
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
bibliography: CaribouDemographicModel.bib
csl: "../templates/journal-of-forest-ecology-and-management.csl" # Insert path for the bib-style
abstract: |
  Across fifty-eight boreal caribou study areas in Canada, both survival and recruitment decrease with the percentage of the study area that is disturbed. There is variation in demographic rates among study areas, particularly where anthropogenic disturbance is low, but where disturbance is high we have no evidence of self-sustaining populations across the country. Demographic projections derived from local population-specific data are  uncertain for populations with limited monitoring. We propose a simple Bayesian population model that integrates prior information from a national analysis of demographic-disturbance relationships with available local demographic data to reduce uncertainty in population viability projections, and to reduce the risk that lack of local data will be used as a reason to delay conservation action. The model also acknowledges additional uncertainty and potential bias due to misidentification of sex or missing calves with a term derived from a simple model of the recruitment survey observation process.  We combine this Bayesian model with simulations of plausible population trajectories in a value of information analysis framework to show how the need for local monitoring varies with landscape condition, and to assess the ability of alternative monitoring scenarios to reduce the risk of errors in population viability projections. Where anthropogenic disturbance is high, reasonably accurate status projections can be made using only national demographic-disturbance relationships. At lower disturbance levels where uncertainty is higher, deploying fewer collars for a moderate number of years may be better than intensive short-term monitoring for improving the reliability of status projections. 
keywords: |
  keyword 1; keyword 2; keyword 3
highlights: |
  These are the highlights. 
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: 72
---

@@

The number of species threatened with extinction is high and getting
higher [@ipbes_global_2019]. Reliable information on the status of
populations and likely impacts of management are needed to inform
conservation action, but effective monitoring is challenging and costly
[@legg_why_2006; @lindenmayer_effective_2018; @wintle_allocating_2010;
@dunham_strategic_2023]. Resources to support conservation and recovery
are limited [@buxton_balancing_2022; @mccarthy_financial_2012;
@gerber_conservation_2016; @wintle_spending_2019], delays in action increase risk
[@martin_acting_2012], and some populations have been monitored to
extinction [@lindenmayer_counting_2013]. A recent analysis of action
plans for Canadian species at risk found that half of tasks are research and
monitoring, and this proportion is highest for species with higher risk
of extinction, highlighting a need to more carefully consider when,
where, and why additional information is needed to inform conservation
action [@buxton_balancing_2022]. Analysis of the implications of
monitoring outcomes for decisions can help to clarify whether additional
monitoring is necessary to inform conservation action
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012].

Analysis methods that can make better use of available monitoring
data to inform decisions include Bayesian population modelling
[@schaub_integrated_2021] and value of information analyses
[@bennett_when_2018; @dunham_strategic_2023; @runge_which_2011;
@gregory_structured_2012]. But to be useful, these methods must be tailored
to particular species and decision contexts. Here, we offer a simple
Bayesian population model for boreal caribou (*Rangifer tarandus
caribou*) that integrates information about the impacts of landscape 
change [@johnson_science_2020] with available demographic data [building on @eacker_web-based_2019] to reduce uncertainty in
population viability projections. Following the value of information analysis
framework proposed by Dunham et al. (2023), we combine this Bayesian
model with simulations of plausible population trajectories
[@stewart_climate-informed_2023; @dyson_existing_2022] to (i) show how the
value of local monitoring varies with landscape condition, and (ii) assess
the ability of alternative monitoring scenarios to reduce the risk of
errors in population viability projections. Our approach is tailored to
boreal caribou, but could be adapted for other species at risk found in
multiple populations that are costly to monitor and sensitive to habitat
disturbance.

Boreal caribou are widely and sparsely distributed through the Canadian
boreal region across a gradient of anthropogenic disturbance
[@eccc_scientific_2011; @johnson_science_2020] that includes some of the
most intact and inaccessible places remaining on earth
[@hirsh-pearson_canadas_2022; @ibisch_global_2016;
@venter_sixteen_2016]. Across 58 boreal caribou study areas, both
survival and recruitment decrease with the percentage of the study area
that is disturbed, and no populations of caribou inhabiting areas with high anthropogenic disturbance are considered self-sustaining [@eccc_scientific_2011; @johnson_science_2020]. The mechanisms behind these
demographic-disturbance relationships include direct impacts on habitat
availability and use, and indirect impacts on predators and alternate
prey [@nbckc_boreal_2022; @eccc_scientific_2011]. Decreases in
recruitment as disturbance increases are also observed over time in
places where there are historic demographic data
[@rudolph_demographic_2017; @hervieux_widespread_2013], but many places
lack sufficient data to characterize local demographic-disturbance
relationships [@johnson_science_2020]. There is also variation in
demographic rates among study areas, particularly where anthropogenic
disturbance is low [@johnson_science_2020; @neufeld_primary_2021;
@fortin_forest_2017; @gagne_impact_2016], and this variation leads to
uncertainty in projections derived from national disturbance-demographic
relationships [@stewart_climate-informed_2023; @dyson_existing_2022].

Interventions such as wolf culls, maternal penning, and restoration of
linear features are intended to improve survival and reproduction of
populations that are not otherwise self-sustaining
[@mcnay_demographic_2022; @serrouya_predicting_2020], and there is a
need to evaluate and compare effects of boreal caribou conservation
actions with careful monitoring. Our focus in this paper is on assessing
and projecting self-sustainability of populations in the absence of
these interventions, to clarify when additional information is needed to
reduce uncertainty about whether there is a need for action. Where
actions are taken, outcomes should be monitored; the model we
propose here does not include effects of management interventions, and
cannot be used to project outcomes for managed populations.

@@

Our proposed model represents an important improvement over less formally integrated methods for estimating and projecting impacts of disturbance on population population viability
[@eccc_scientific_2011; @johnson_science_2020] because it reconciles and
accounts for uncertainties in local demographic data and
demographic-disturbance relationships. We show that integrating the most
informative predictor of outcomes for boreal caribou across their range
can substantially reduce uncertainty in local demographic projections
and status assessments, can improve viability projections on changing
landscapes, and can help clarify where, why and when additional data is
required to inform decisions and project outcomes on changing
landscapes. We hope that this approach will help to identify where available
information is sufficient for assessing and projecting population
viability, and thereby reduce the risk that uncertainty is used as a
reason to delay conservation action for imperiled populations.

@@

To derive a composition survey bias term we assume each group
of animals in a calf:cow composition survey contains one or more
collared adult females ($T$), and may also include: uncollared adult
females misidentified as young bulls or unknown sex ($U$); correctly
identified uncollared adult females ($V$); young bulls correctly
identified as male or unknown sex ($O$); young bulls misidentified as
uncollared adult females ($P$); observed calves ($J$); and unobserved
calves ($K$). The apparent number of adult females in the group is
$T+V+P=Tw$, where $w$ is the observed ratio of the apparent number of
uncollared adult females to collared animals in the recruitment survey.
The ratio of young bulls to uncollared adult females in the group is
$q = (P+O)/(U+V).$ Assuming an equal probability $u$ of misidentifying
young bulls as adult females and vice versa, we get $V=(U+V)(1-u)$ and
$P=(O+P)u$. Given a probability $z$ of missing calves, we get
$J=(J+K)(1-z)$. Given these assumptions and definitions, the sex and
bias-corrected recruitment rate $X$ can be written as a function of the
observed calf:cow ratio $R$, the cow multiplier $w$, the ratio of young
bulls to adult females $q$, and the misidentification probabilities $u$
and $z$ (see Supplement B for details). For simplicity, we combine the
probabilities into a bias term $c$ that is integrated into
the Bayesian population model and simulations (below):

\begin{equation}
  c=\frac{w(1+qu-u)}{(w+qu-u)(1-z)}; X=cR/2.
  (\#eq:c)
\end{equation}

@@

In simulated monitoring scenarios, we assume collars are deployed in January, and
continue to function for up to 6 years. The target number of collared females $n$ varies
among scenarios, as does the total duration of the monitoring program
$d$ and the ratio of uncollared to collared adults in the recruitment
survey $w$ (Table \@ref(tab:monitoringScenarios)). Collars lost to
mortality may be replaced each year ($o=1$), or there may be a 3 year
gap between deployments ($o=4$).

@@

$\phi^R \sim \text{Normal}(19.862,2.229)$ and
$\phi^S \sim \text{Normal}(63.733,8.311)$ are precisions of the Beta
distributed errors [@ferrari_beta_2004]. At the beginning of a
simulation for an example population, regression coefficient values are
sampled from Gaussian distributions (see Table \@ref(tab:priors)) and
the population is assigned to quantiles of the Beta error distributions
for survival and recruitment. The population remains in these quantiles
as disturbance changes over time, so there is substantial persistent
variation in recruitment and survival among example populations (Figure
\@ref(fig:exampleTrajectories)). To ensure that results are comparable across
monitoring scenarios and that populations remain extant through the
monitoring period, all simulations are started with $N_0=5000$ animals
24 years prior to the start of the projection period. Realized population growth rate is $\dot{\lambda}_t=\dot{N}_t/\dot{N}_{t-1}$ Expected population growth rate (without interannual variation, demographic stochasticity, or effects of population density) is $\bar{\lambda}_t=\bar{S}_t(1+\dot{c}\bar{R}_t/2)$.

```{r exampleTrajectories, fig.cap="Examples of simulated population trajectories from a model informed only by national disturbance-demographic relationships. To highlight how variation among trajectories persists as disturbance changes, we have reduced interannual variation by 75\\% in these examples ($v_R=0.115,v_S=0.02175$). Orange dots show means of 500 example trajectories. For reference, black lines and and shaded bands show means and 95\\% predictive intervals from Johnson et al's (2020) regression models.", cache=TRUE}

@@

As done by Johnson et al. [-@johnson_science_2020], we model survival
probability as an adjusted function of buffered anthropogenic
disturbance $A_t$ with a log link and Beta distributed interannual variation with coefficient of variation $\nu_S$, mean $\tilde{S_t}$ and standard deviation $\sigma_{S,t}$ [@stewart_climate-informed_2023;
@dyson_existing_2022]: 
\begin{equation}
\tilde{S_t}=(46 e^{\beta^S_0+\beta^S_a A_t}-0.5)/45
  (\#eq:Stilde)
\end{equation}

@@

For reference, all variables and parameters are defined in supplemental Table A1. Models and analysis were implemented using the R language
[@r_core_team_r_2023] and a variety of *tidyverse* packages
[@wickham_welcome_2019]. We use the open source Gibbs sampling software
JAGS [@plummer_jags_2023] and the *R2jags* R package [@su_r2jags_2021]
to fit the Bayesian model. We ran 15,000 iterations of 4 chains,
discarded the first 10,000 interations, and used a thinning rate of 2.
We assessed convergence for a small random subset of example
trajectories using a threshold for the Gelman-Rubin diagnostic
[@gelman_inference_1992] of r-hat \< 1.1, and visually inspected
posterior autocorrelation and traceplots produced by the *mcmcplots* R
package [@curtis_mcmcplots_2018]. Code required to run analyses and
reproduce figures in this paper is available at
<https://github.com/LandSciTech/Caribou-Demographic-Projection-Paper>,
and in the *caribouMetrics* R package available on GitHub at
<https://github.com/LandSciTech/caribouMetrics>.

@@

Analysis of simulated example demographic trajectories confirms that our
Bayesian model can effectively combine local demographic data with
priors from a national demographic-disturbance relationships to reduce
uncertainty in boreal caribou demographic projections (Figures
\@ref(fig:examplePredictions) and \@ref(fig:figDiffs)). As expected,
uncertainty about the population growth rate decreases as sampling
effort in each year (i.e. number of collars) or the number of years of
monitoring (Figure \@ref(fig:figDiffs)) increase. Reducing uncertainty is more challenging when interannual variation is high (Figure
\@ref(fig:examplePredictions)). Increasing the forecast horizon [@petchey_ecological_2015] does not increase uncertainty about expected population growth (compare 5 and 20 year projections in
Figure \@ref(fig:figDiffs)) in these scenarios.

@@

The availability of boreal caribou demographic data varies widely across
Canada due to large variation in habitat condition, population
condition, available resources for monitoring, and remoteness. In a few
regions enough data have been collected over time to characterize local
demographic-disturbance relationships [@rudolph_demographic_2017;
@hervieux_widespread_2013], but most populations have only been sparsely
monitored for snapshot estimates of trend or abundance
[@johnson_science_2020]. As a result, in most regions it is necessary to
make assumptions informed by observed variation in demographic rates
with disturbance among populations and other relevant information about
impacts of disturbance on caribou. However, high observed variation in
demography where anthropogenic disturbance is low leads to high
uncertainty in projections informed by national demographic-disturbance
relationships alone [@stewart_climate-informed_2023;
@dyson_existing_2022]. In some cases, predictions of population growth
rate informed by national demographic-disturbance relationships appear
to disagree with estimates from local demographic data
[@eccc_scientific_2011; @dyson_existing_2022], but without formal
methods of accounting for uncertainty it can be difficult to distinguish
true discrepancies from stochastic variation.

We have proposed a simple Bayesian population model that helps address these challenges by
acknowledging uncertainty associated with local demographic data and
national demographic-disturbance relationships, and integrating these
two sources of information to reduce uncertainty in projections of the
impacts of changing landscape conditions. This Bayesian approach
represents an important improvement over less formally integrated
methods for estimating and projecting impacts of disturbance on
population population viability [@eccc_scientific_2011;
@johnson_science_2020] because it reconciles and accounts for
uncertainties in local demographic data and in demographic-disturbance
relationships.

Our model assumes that increasing cumulative anthropogenic disturbance
causes a log-linear decrease in caribou demographic rates (equations 4,
5, 10 and 13), reflecting a scientific consensus that cumulative anthropogenic disturbance has a
negative impact on boreal caribou [@eccc_scientific_2011;
@johnson_science_2020; @gouvernement_du_quebec_revue_2021;
@rudolph_demographic_2017; @sorensen_determining_2008;
@hervieux_widespread_2013]. Local data can be integrated to estimate
local demographic-disturbance parameters that differ from the national
means, and in the absence of local data, appropriate priors lead to a
range of outcomes consistent with the observed range of variation across
the country [Figures 3 and 5 in @johnson_science_2020]. Others
[@moeller_integrating_2021; @mcnay_demographic_2022;
@eacker_web-based_2019; @dalgarno_bboutools_2024] have also used Bayesian models to estimate
caribou demographic rates from local monitoring data, but our addition
of prior information from national demographic-disturbance relationships
is novel, and important for reducing the risk that lack of local data
will be used as a reason to delay conservation action for highly
disturbed and poorly monitored populations. Integrating
demographic-disturbance relationships into a viability projection model
clarifies that no additional information is needed to project outcomes
where disturbance is high and actions to support recovery are not taken.

@@

Variation in anthropogenic disturbance across the Canadian boreal region
is partially correlated with variation in climate and primary
productivity [@neilson_direct_2022] because forestry occurs in higher
productivity areas, and petroleum extraction activities are concentrated
in Western Canada. Effects of disturbance on vegetation and foodweb
dynamics vary with primary productivity [@neufeld_primary_2021], and
some have expressed concern about confounding of demographic-disturbance
relationships [@wilson_causal_2021]. One option for addressing these
concerns would be to develop regional demographic-disturbance
relationships for populations that are ecologically similar to one
another (e.g. caribou inhabiting forests that are sufficiently
productive for industrial forestry in Eastern Canada)
[@eccc_scientific_2024], and to use the priors from regional
demographic-disturbance relationships to inform demographic projections
within the region. However, while acknowledging and allowing for
regional variation, we also need to guard against risks that uncertainty
will be inflated and used as an excuse to delay action, a tactic that
has been used in this and other decision contexts
[@bissonnette_controverses_2020; @boan_climate_2018]. In order to be
informative and credible, regional demographic-disturbance analyses must
include a sufficient number of observations spanning a range from high
to low anthropogenic disturbance. Priors that lead to projections of
outcomes for caribou that fall outside the observed range of variation
across the country, characterized here (Figure 2) by 95% prediction
intervals from a national analysis (Johnson et al. 2020), are
implausible. In other words, if we have seen no examples of
self-sustaining caribou populations at a particularly high level of
disturbance, it is unreasonable to project or expect this outcome. These
premises constrain the plausible demographic-disturbance priors, and
clarify that no additional information is needed to predict outcomes for
caribou where anthropogenic disturbance is high (Figure 5) and no
meaningful actions are taken to support recovery.

@@

Another novel aspect of this work is a term that acknowledges possible bias
caused by misclassification of adults and missing calves in
composition surveys [@ellington_divergent_2020;
@decesare_estimating_2012]. Lacking information about the true magnitude
of these errors, we examined sensitivity to a range of assumed values, and
found that while acknowledging bias did increase uncertainty, these
increases were not large (Supplement E). A key reason for this result is
that we assumed the same range of values for the probability of
misclassifying adults and the probability of missing calves. Presuming
that surveyed groups contain more cows than bulls, and that there is an
equal probability of misclassifying cows as young bulls and vice versa,
then misclassification errors cause high estimated recruitment. Missing
calves causes low estimated recruitment, and together these errors tend
to cancel one another out. Given lack of knowledge about true composition survey error rates, the most appropriate interpretation of our results is that better information is required in order to assess the importance of these errors.
